\section{Introduction}
% Argues: production LLM traffic is non-stationary, offline-frozen static configs lose 9-10% throughput and ~8x TTFT p99 under regime shift, and the central thesis that adaptive online scheduling (not speculative decoding) is the driver; states contributions.

Large language model (LLM) serving systems are configured offline. An
operator selects a representative trace, grid-searches over the
scheduler's knobs --- maximum batch size, chunked-prefill size,
speculative-decoding depth, KV-cache eviction policy --- and freezes the
single configuration that maximizes validation throughput. This static
configuration then serves all subsequent traffic. The premise is that a
good operating point, once found, remains good. For a fixed workload it
does.

Production traffic, however, is not fixed. It is non-stationary: it
alternates between latency-sensitive \emph{interactive} bursts and
throughput-oriented \emph{batch} phases, and the two regimes have nearly
disjoint optima. In our measurements the best interactive configuration
uses a small batch (\texttt{max\_batch\_size=4}), while the best batch
configuration uses a large batch (\texttt{max\_batch\_size=32}) with
preemption disabled. No single static point can
be both. A configuration tuned for aggregate throughput is therefore
mis-sized for whichever regime it was not tuned for, and the cost is
large rather than marginal. On non-stationary traffic a per-model-tuned
static-best configuration sacrifices 9--10\% throughput against an
adaptive policy, and its tail latency collapses: on TinyLlama the
static-best TTFT p99 inflates by roughly $8\times$, from 2,199~ms to
18,058~ms, because batch-sized batching settings starve interactive
requests the moment the regime flips.

The natural response is to tune online. Yet naive reactive autotuning
makes the problem worse, not better. A hill-climbing autotuner that
chases instantaneous reward without a model of the regime structure
thrashes: it perturbs the configuration continuously, pays exploration
cost on every step, and never recognizes that the workload has a small
number of stable modes. Measured against a per-regime oracle, our
\textsc{AutoTuner} baseline leaves an 18.9\% mean throughput gap ---
\emph{wider} than the 16.2\% gap of the frozen static configuration it
was meant to improve upon. Reacting to the wrong signal is worse than
not reacting at all.

The surprising finding of this paper is that the benefit of adaptation
is \emph{entirely conditional on regime change}. On every stationary,
single-regime workload we test --- short prompts, long prompts,
interactive-only, batch-only --- our adaptive system is not merely close
to the static-best but statistically indistinguishable from it: the mean
throughput difference is exactly zero, with Cohen's $d = 0$ and
$p = 1.0$. The entire advantage materializes only under
non-stationarity, where it is enormous (paired $d = 23.2$,
$p = 2.3\times10^{-41}$ over $n=30$ seeds). This dichotomy is
diagnostic. Because the serving engine --- paged attention, continuous
batching, and a fixed, disabled speculative-decoding setting --- is
identical across both the stationary and the non-stationary
conditions, the source of the gain cannot be any static mechanism. The
only thing that differs between the conditions is whether the operating
point is re-selected as the workload shifts. We conclude that the driver
of performance under workload change is online re-selection of the
configuration itself --- adaptive scheduling --- and not the underlying
serving mechanisms that a static system already provides.

We make this concrete with CARL, a contextual-bandit scheduler that
treats each candidate serving configuration as an arm and re-selects the
operating point online as the regime shifts. CARL observes lightweight
workload features, applies a LinUCB contextual-bandit policy to choose a
configuration, and reconfigures the unmodified serving engine
accordingly; no part of the inference path is changed. Across two model
families --- TinyLlama-1.1B and Qwen2-0.5B --- and with zero per-model
retuning, CARL reaches $108.2 \pm 3.4\%$ of each model's own
per-model-tuned static-best (+10.6\% on TinyLlama, +5.8\% on Qwen2) on
real-GPU non-stationary traffic. Its LinUCB policy closes the per-regime
oracle gap from the static configuration's 16.2\% to 0.45\% --- within
noise of the oracle that knows the best arm for each regime in advance.
The adaptation is well-behaved: CARL converges to the per-regime optimum
deterministically at cycle~9 across all seeds, with a total cumulative
regret of 2.583 against the oracle.

We are equally explicit about where CARL does \emph{not} help. On stress
workloads designed to expose it --- stable load, single-queue
sequential serving, rapid oscillation, and memory pressure --- CARL ties
or loses in all five cases, by as much as 24.2\% on rapid oscillation,
where the regime changes faster than the bandit can amortize its
exploration. This is not a caveat to be hidden but the defining boundary
of the method: adaptation is a net benefit precisely when the workload
is non-stationary at a timescale long enough to learn, and a net cost
otherwise. Regime non-stationarity is both the problem CARL solves and
the precondition for using it.

\paragraph{Contributions.}
\begin{itemize}
  \item We identify \emph{regime non-stationarity} as the root cause of
    static-serving suboptimality, showing that the interactive and batch
    regimes have nearly disjoint optima and that a single frozen
    configuration loses 9--10\% throughput and $\sim$$8\times$ TTFT p99
    under regime shift.
  \item We show that adaptive online scheduling, rather than speculative
    decoding or any other static serving mechanism, is the primary
    driver of performance under workload change: the gain is zero on
    stationary workloads ($d=0$, $p=1.0$) and appears only under regime
    change ($d=23.2$, $p=2.3\times10^{-41}$) with the engine held fixed.
  \item We present CARL, a contextual-bandit (LinUCB) controller that
    re-selects the serving configuration per regime over an unmodified
    engine and closes the per-regime oracle gap from 16.2\% to 0.45\%,
    reaching 108.2\% of per-model-tuned static-best across two model
    families with zero retuning.
  \item We give a convergence analysis showing deterministic convergence
    to the per-regime optimum (cycle~9 across all seeds) with bounded
    cumulative regret (2.583 against the oracle).
  \item We provide an honest failure-case characterization that bounds
    the claim, showing CARL ties or loses on all five stationary or
    fast-oscillating stress workloads (up to $-$24.2\%), and use it to
    state the precise operating envelope of adaptive serving.
\end{itemize}
