\section{Evaluation}
% Argues: experimental setup and headline results --- real-GPU cross-model gains (108.2% of static-best, +10.6% TinyLlama, +5.8% Qwen2), throughput/TTFT wins on non-stationary traffic, and statistical validation (n=30 paired t-test, large effect sizes).

\subsection{Design Scope and Limitations}
\label{sec:limitations}

Speculative decoding in our implementation is single-request only: the
scheduler runs the greedy self-speculative (early-exit) path solely when
exactly one request is in the decode phase and reverts to standard
batched decoding whenever two or more requests decode concurrently, and
the implemented draft/target decoder is not wired into the paged-cache
scheduler --- so speculation is inactive across the batched regime and is
held disabled in the experiments reported here.

We quantify why speculation is disabled rather than merely descoped. On a
single Tesla~T4, we swept the engine's self-speculative early-exit decoder
over draft depth (exit layer $L\in\{4,6,8,10,12\}$) and draft length
$K\in\{1,\dots,8\}$, serving one request at a time so the self-speculative
path is active, with CUDA graphs off for both arms for an apples-to-apples
comparison (Table~\ref{tab:specbreakeven}). Draft acceptance is low across
the entire grid: it peaks at only $5.1\%$ (at $L{=}10$, $K{=}1$) and is
effectively zero for the shallowest exits ($L{\le}6$). An analytical
break-even model --- the per-token acceptance at which one speculative round
recovers its own draft cost --- requires $19\%$--$83\%$ acceptance depending
on $(L,K)$; even at the single most favorable operating point, the measured
$5.1\%$ is nearly an order of magnitude below the $42.5\%$ needed just to
break even there. Consistently, every one of the $25$ configurations is a net
slowdown ($0.21\times$--$0.84\times$ the vanilla decode rate); none reaches
break-even. This is why self-speculation contributes nothing on our
model/hardware and is held disabled in the main experiments: at this model
scale the early-exit draft is too weak to pay for the extra forward pass.

We report this as a \emph{single-seed} characterization (seed~42, one T4 run,
three prompts per cell) intended to establish the \emph{scope} of the design,
not a statistically powered measurement. We therefore make no significance
claim and attach no confidence intervals; the gap between achieved and
required acceptance is qualitative and nearly an order of magnitude, which is
what supports the design decision. Although this characterization uses a
single seed, the measured acceptance rate remains nearly an order of
magnitude below the analytical break-even threshold, making the qualitative
conclusion unlikely to change under additional seeds.

\begin{table}[t]
\centering
\small
\caption{Self-speculative decode speedup vs.\ vanilla by exit layer $L$ and
draft length $K$ (TinyLlama-1.1B, single Tesla~T4, seed~42, three prompts per
cell). All values $<1$: speculation is a net slowdown at every setting; none
reaches break-even. Single-seed characterization.}
\label{tab:specbreakeven}
\begin{tabular}{lrrrrr}
\toprule
$L \backslash K$ & $K{=}1$ & $K{=}2$ & $K{=}4$ & $K{=}6$ & $K{=}8$ \\
\midrule
$L{=}4$  & 0.84 & 0.73 & 0.57 & 0.48 & 0.41 \\
$L{=}6$  & 0.79 & 0.64 & 0.49 & 0.39 & 0.33 \\
$L{=}8$  & 0.77 & 0.62 & 0.44 & 0.35 & 0.29 \\
$L{=}10$ & 0.74 & 0.57 & 0.39 & 0.31 & 0.25 \\
$L{=}12$ & 0.67 & 0.50 & 0.34 & 0.26 & 0.21 \\
\bottomrule
\end{tabular}
\end{table}
