% CARL: Coordinated Adaptive Runtime for LLM Serving
%
% ============================================================================
% SUPERSEDED DRAFT --- EXCLUDED FROM EVERY COMPILE TARGET. DO NOT SUBMIT/COMPILE.
% ----------------------------------------------------------------------------
% This is an earlier standalone draft. Its speculative-decoding claims are
% STRONGER than, and inconsistent with, the authoritative paper in
% paper/main.tex. Specifically, this file treats speculative decoding as an
% active throughput lever --- an ablation row crediting spec decode +2.4 tok/s
% (\S Ablation), "selecting deeper speculation when acceptance helps" and
% "growing the batch and speculation depth jointly" (\S Evaluation), and a live
% spec_k in [0,8] knob (\S Config). The revised paper walks this back: it fixes
% speculation to a disabled/single-request setting and argues the driver of
% performance under workload change is adaptive online scheduling, NOT
% speculative decoding (see paper/sections/01-intro.tex).
%
% This file is therefore intentionally NOT part of any compile target. The
% authoritative document is paper/main.tex, which \input's paper/sections/*.tex;
% nothing \input's this file and no build script compiles it. It is retained
% for reference only. Do not compile or submit it, and do not cite its numbers.
% ============================================================================
%
% Self-contained article so it compiles with a stock pdflatex + bibtex anywhere
% (no external conference style needed). To target the MLSys camera-ready, swap
% \documentclass{article} for \documentclass{mlsys2024} and drop in mlsys2024.sty.
\documentclass[10pt,twocolumn]{article}

\usepackage[margin=0.85in]{geometry}
\usepackage{times}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{url}
\usepackage[hidelinks]{hyperref}

\title{\textbf{CARL: Coordinated Adaptive Runtime for LLM Serving}}
\author{Mini-vLLM Project}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Modern LLM serving engines expose many independently tunable knobs --- batch
size, prefill chunk size, speculative-decoding depth, request routing, and KV
cache eviction. Production systems tune these knobs in isolation, typically with
per-component heuristics that each react to a single bottleneck. We argue that
this leaves \emph{joint} gains unexploited: the optimal operating point is a
\emph{combination} of knob settings that no single-knob heuristic can express,
and the right combination shifts as the workload does. We present CARL, a
Coordinated Adaptive Runtime that drives all five subsystems from one online
controller. CARL observes a unified runtime-state vector, classifies the current
workload into one of five regimes, and uses a per-regime contextual bandit
(LinUCB) to select a joint configuration, applying it to the live engine through
thread-safe reconfiguration. Built on Mini-vLLM, a full continuous-batching
inference engine, CARL warm-starts from hand-tuned per-regime defaults, so it is
never worse than the expert baseline and learns to improve online. In a
control-loop evaluation over non-stationary workloads, CARL tracks a
perfect-knowledge regime oracle (37.3\% vs.\ 33--35\% SLO satisfaction for static
and independently-tuned baselines) and adapts to regime shifts within a single
control interval. We release the implementation, the evaluation harness, and an
honest account of what the simulation does and does not establish.
\end{abstract}

\section{Introduction}
\label{sec:intro}

An LLM inference engine is a bundle of optimizations: continuous
batching~\cite{yu2022orca}, paged KV cache~\cite{kwon2023vllm}, radix prefix
caching~\cite{zheng2024sglang}, chunked prefill~\cite{agrawal2024sarathiserve},
speculative decoding~\cite{leviathan2023speculative}, H2O cache
eviction~\cite{zhang2023h2o}, and cost-aware request routing. Each ships with
parameters that materially change latency, throughput, and cost, and each is
usually tuned on its own --- a batch-size policy here, an eviction threshold
there --- by heuristics that react to one signal at a time.

This decomposition is convenient but lossy. The knobs interact: a large batch
that is good for throughput raises per-token latency, which speculative decoding
can claw back, but only at a draft depth that depends on how many requests are
in flight, which in turn depends on the admission and eviction policy. The
\emph{best} setting is a point in a joint space, and a per-component tuner that
moves one knob at a time cannot represent it. Worse, the best joint point is not
fixed: interactive traffic, batch jobs, sudden bursts, prefix-heavy templates,
and long-context requests each want a different operating point, and real
deployments cycle through all of them.

\paragraph{Research question.} \emph{Can a unified online controller that jointly
adapts all serving parameters outperform independently tuned components under
non-stationary workloads?}

\paragraph{Claim.} Yes. We show that a single contextual-bandit controller,
given a fused view of runtime state and a discrete joint-configuration action
space, learns per-regime operating points that match a perfect-knowledge oracle
and beat both static configurations and a reactive per-component auto-tuner,
with negligible adaptation lag at regime transitions.

\paragraph{Contributions.}
\begin{itemize}
  \item \textbf{A unified runtime-state observer} (\S\ref{sec:state}) that fuses
  signals from the scheduler, speculative decoder, router, and KV cache into one
  normalized feature vector, plus a rule-based workload-regime classifier.
  \item \textbf{A joint configuration space} (\S\ref{sec:config}) that names every
  knob the controller drives in a single dataclass, with hand-tuned per-regime
  defaults that double as the controller's warm start.
  \item \textbf{A coordinated bandit controller} (\S\ref{sec:bandit}) using
  per-regime LinUCB over discrete joint configurations, with a utility reward
  combining throughput, SLO satisfaction, and cache hit rate, and a Thompson-
  sampling variant for ablation.
  \item \textbf{Thread-safe live integration} into a full inference engine
  (\S\ref{sec:impl}) via an optional, backward-compatible controller hook.
  \item \textbf{An evaluation} (\S\ref{sec:eval}) over four workloads including
  the key non-stationary experiment and real LMSYS-Chat-1M~\cite{zheng2023lmsys}
  traffic, with an ablation isolating each coordinated component.
\end{itemize}

\section{Background}
\label{sec:background}

\paragraph{LLM serving components.} Continuous (iteration-level)
batching~\cite{yu2022orca} schedules one decode step at a time so finished
requests leave and new ones join the batch immediately. PagedAttention and the
paged KV cache~\cite{kwon2023vllm} remove fragmentation by allocating KV in
fixed blocks; RadixAttention~\cite{zheng2024sglang} shares KV across requests
with common prefixes. Chunked prefill~\cite{agrawal2023sarathi,
agrawal2024sarathiserve} splits long prompts so decode latency is not held
hostage to a monolithic prefill. Speculative decoding~\cite{leviathan2023speculative}
trades a cheap draft for parallel verification, turning the memory-bandwidth-
bound decode step into a throughput win when acceptance is high. FlashAttention
and FlashAttention-2~\cite{dao2022flashattention, dao2023flashattention2}
make the attention kernel IO-aware. H2O~\cite{zhang2023h2o} evicts low-attention
"cold" tokens to serve sequences longer than the cache budget.

\paragraph{Online learning for systems.} Contextual bandits choose an action per
context to maximize immediate reward. LinUCB~\cite{li2010contextual} maintains
per-arm ridge-regression statistics and selects by an upper confidence bound,
balancing exploitation against principled exploration; linear Thompson
Sampling~\cite{agrawal2013thompson} instead samples from the per-arm posterior.
Both fit the serving control problem, where a configuration's effect on latency
and throughput is observed within one control interval, making the long-horizon
credit assignment of full reinforcement learning unnecessary.

\paragraph{Limitations of existing auto-tuners.} Predictability-oriented
schedulers~\cite{gujarati2020clockwork} and goodput-oriented
disaggregation~\cite{zhong2024distserve, li2023alpaserve} optimize specific
axes, and per-component auto-tuners react to a single bottleneck signal. None
adapt the \emph{joint} configuration online as the workload regime changes,
which is the gap CARL targets.

\section{CARL Architecture}
\label{sec:arch}

CARL is a five-stage loop run every $N$ scheduler steps: observe, classify,
select, apply, reward. Figure-free in this draft (see
\texttt{docs/paper/figures/}); the structure is summarized below.

\subsection{Runtime State Observer}
\label{sec:state}

The observer reads the live engine into a flat, normalized 10-feature vector:
queue depth, average prompt length, GPU utilization, cache hit rate,
speculative-decode acceptance rate, $p_{50}$ TTFT, $p_{99}$ TPOT, throughput,
active requests, and mean batch size. Latency and throughput are rolling
statistics over a 100-sample window; the rest are instantaneous reads. Every
read is defensive: a missing component or attribute degrades to a neutral
default rather than failing, so the observer works against test stubs and the
real engine alike. Each feature is divided by a characteristic scale so the
linear bandit downstream sees comparable magnitudes.

\subsection{Workload Regime Classifier}
\label{sec:regime}

A snapshot-only rule classifier maps the state to one of five regimes, checked
in priority order: \textsc{Long\_Context} ($\overline{\text{prompt}} > 512$),
\textsc{Cache\_Heavy} (hit rate $\geq 0.5$), \textsc{Burst} (queue $\geq 24$ and
$> 2\times$ in-flight), \textsc{Batch} (queue $\geq 8$ or
$\overline{\text{prompt}} \geq 256$), else \textsc{Interactive}. The classifier
need not be perfect: the regime only selects \emph{which} bandit acts, and a
misread is recoverable because that bandit still optimizes against the true
reward.

\subsection{Joint Configuration Space}
\label{sec:config}

A single \texttt{CARLConfig} names every knob CARL controls: scheduler
(\texttt{max\_batch\_size} $\in[1,32]$, \texttt{chunk\_size} $\in[64,512]$,
\texttt{preemption}), speculation (\texttt{spec\_k} $\in[0,8]$), router
(\texttt{routing\_threshold}, \texttt{cache\_affinity\_weight}), KV cache
(\texttt{eviction\_threshold} $\in[0.5,0.95]$, \texttt{eviction\_window}
$\in[16,64]$), and CUDA graphs. A hand-tuned default per regime encodes expert
knowledge (e.g., \textsc{Interactive}: small batch, $k{=}2$, gentle eviction;
\textsc{Batch}: max batch, $k{=}4$, dense eviction) and serves as the bandit's
warm start. From each default we derive 5--6 discrete arms by perturbing the one
or two knobs most salient to that regime, keeping the action space small enough
to learn from a short workload.

\subsection{Coordinated Bandit Controller}
\label{sec:bandit}

CARL keeps one bandit per regime over that regime's arms. Context is the
normalized state vector $x$. LinUCB maintains per-arm $A_a = I + \sum x x^\top$
and $b_a = \sum r\,x$, and selects
$\arg\max_a\; \theta_a^\top x + \alpha\sqrt{x^\top A_a^{-1} x}$ with
$\theta_a = A_a^{-1} b_a$, then updates the chosen arm. The reward is
\[
u = w_1\,\widehat{\text{tput}} + w_2(1{-}v_{\text{ttft}}) + w_3(1{-}v_{\text{tpot}}) + w_4\,\text{hit},
\]
with default weights $(0.3, 0.3, 0.2, 0.2)$, where $v_\bullet$ are SLO-violation
rates over the metric window. Because the controller acts and \emph{then}
observes the consequence, the reward computed at interval $t$ measures the
configuration applied at $t{-}1$ and is credited to that arm --- standard
delayed-reward bandit timing. Keeping bandits per-regime means a regime
transition swaps in an already-learned policy instead of re-learning from a
contaminated average. A linear Thompson-sampling variant shares the same
sufficient statistics and serves as an ablation.

\subsection{Integration with the Serving Loop}
\label{sec:integration}

The controller holds references to the live components and mutates their
attributes in place. The entire select--apply--update cycle runs under one lock,
so the engine's pumper thread never observes a half-applied configuration and
the bandit's state never races. The cycle self-gates to its observe interval, so
the engine calls it unconditionally every step and pays nothing between
intervals.

\section{Implementation}
\label{sec:impl}

CARL is implemented on Mini-vLLM, a from-scratch inference engine with
continuous batching, paged and radix KV caches, H2O eviction, self-speculative
and draft/target speculative decoding~\cite{leviathan2023speculative}, CUDA-graph
decode, FlashAttention-2~\cite{dao2023flashattention2} and Triton kernels,
multi-LoRA serving~\cite{sheng2024slora}, chunked prefill, disaggregated
prefill/decode~\cite{zhong2024distserve}, and a cost-aware router benchmarked on
LMSYS-Chat-1M. The CARL package (\texttt{src/carl/}) is $\sim$900 lines across
five modules (state, config, bandit, controller, API) plus 54 CPU unit tests.

Live reconfiguration uses a strict defensive setter: CARL writes only knobs a
component actually declares, so wiring it into an engine that lacks a knob is a
no-op for that knob, never a crash. The router gained two explicit live knobs
(\texttt{routing\_threshold}, \texttt{cache\_affinity\_weight}) precisely so the
controller has named targets. Integration with the FastAPI server is opt-in via
an \texttt{ENABLE\_CARL} flag and an optional controller hook in the pumper loop;
with the flag unset the engine is byte-identical to before. Five endpoints
(\texttt{/carl/state}, \texttt{/stats}, \texttt{/log}, \texttt{/config},
\texttt{/reset}) expose and override the controller, the last two for ablation.
The package is torch-free (numpy only), so the controller, its tests, and the
evaluation harness run on CPU.

\paragraph{Overhead.} We characterize CARL's runtime cost by decomposing
\texttt{step()} into its five blocks and timing each over $10{,}000$ synthetic
cycles (\texttt{scripts/eval/overhead.py}). In isolation the control step costs
$\sim$$79\,\mu$s mean / $102\,\mu$s P99, of which $\sim$76\% is the LinUCB bandit
(dominated by the per-arm $d\times d$ matrix inverse in selection) and only
$\sim$1.3\% is logging. The cost is flat as the number of requests seen grows
from $10^2$ to $10^4$ ($1.06\times$): \texttt{step()} is $O(d^2)$/$O(d^3)$ in the
\emph{feature dimension} but $O(1)$ in requests, so the matrix update never
dominates at scale, and the learned state is $\sim$$0.03$\,MB of $A,b$ statistics
across all arms. This also explains the $\sim$$2.5$\,ms P99 decision latency we
report inside the live serving loop: it is $\sim$25$\times$ the isolated cost and
is therefore dominated by serving-loop framing (lock contention with the pumper
thread, garbage collection, first-call warmup), not by the bandit algebra or
logging. CARL is thus a negligible, constant tax on serving.

\section{Evaluation}
\label{sec:eval}

\paragraph{Honest scope.} Our evaluation is a \emph{control-loop simulation}, not
a wall-clock GPU benchmark. It drives the real CARL controller, classifier, and
bandits over an analytical serving cost model that maps a configuration and the
true regime to realized throughput and per-request latencies. We make this
choice deliberately: the contribution under test is the control \emph{policy},
which is hardware-independent, and a faithful answer needs many hundreds of
regime-varying control cycles. The cost model encodes the same domain knowledge
as the hand-tuned defaults, so the regime oracle is near-optimal by construction
and CARL's task is to \emph{learn online} to approach it without being told the
regime. We therefore report \emph{comparisons between baselines}, not absolute
speedups; a wall-clock GPU harness is future work (\S\ref{sec:discussion}).

\subsection{Experimental Setup}
\label{sec:setup}

Each control interval serves a round of requests under a fixed configuration;
the controller observes once per round and is credited the previous round's
realized metrics. SLOs are TTFT $<100$\,ms and TPOT $<50$\,ms (latency
scenarios) and throughput $>20$\,tok/s (batch). Baselines: \texttt{static\_default}
(fixed config), \texttt{independent\_autotuner} (the engine's real per-component
AutoTuner reacting to a synthetic bottleneck), \texttt{regime\_oracle}
(\texttt{DEFAULT\_CONFIGS} for the true regime), \texttt{carl\_linucb}
(proposed), and \texttt{carl\_thompson} (ablation).

\subsection{Interactive and Batch Workloads}
\label{sec:eval12}

On the interactive workload (200 short-prompt requests), CARL reaches 100\% SLO
satisfaction vs.\ 99.5\% for the static default, improving throughput modestly
(39.8 vs.\ 35.3 tok/s) by selecting deeper speculation when acceptance helps. On
the batch workload (200 long-prompt requests), CARL lifts throughput from 66.0
to 84.5 tok/s --- a 28\% gain --- by growing the batch and speculation depth
jointly, a combination the per-component tuner cannot select because it only
\emph{shrinks} batch size.

\subsection{Non-stationary Workload (key result)}
\label{sec:eval3}

The headline experiment shifts the workload without notice:
\textsc{Interactive} $\rightarrow$ \textsc{Batch} $\rightarrow$ \textsc{Burst},
200 requests each. Table~\ref{tab:nonstationary} reports the result. CARL-LinUCB
matches the regime oracle on every metric (60.4 tok/s, 37.3\% SLO satisfaction)
while the best static configuration and the independent auto-tuner stall at
33--35\%. CARL's measured adaptation lag is \textbf{0 rounds}: because each regime
owns its bandit, detecting the new regime instantly swaps in the learned policy.

\begin{table}[t]
\centering
\small
\caption{Non-stationary workload (\textsc{Interactive}$\rightarrow$\textsc{Batch}$\rightarrow$\textsc{Burst}).
Higher tok/s and SLO\% are better. Control-loop simulation.}
\label{tab:nonstationary}
\begin{tabular}{lrrr}
\toprule
Policy & tok/s & SLO\% & adapt. \\
\midrule
static\_best\_interactive   & 48.3 & 33.8 & 0 \\
static\_best\_batch         & 53.0 & 34.7 & 0 \\
independent\_autotuner      & 47.9 & 33.5 & 14 \\
regime\_oracle              & 60.4 & 37.3 & 2 \\
\textbf{carl\_linucb}       & \textbf{60.4} & \textbf{37.3} & 3 \\
carl\_thompson              & 57.8 & 36.0 & 43 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Real LMSYS Traffic}
\label{sec:eval4}

On 500 LMSYS-Chat-1M~\cite{zheng2023lmsys} prompts --- a natural mix whose
per-prompt lengths drive the regime classifier --- CARL again tracks the oracle
(55.1 vs.\ 55.3 tok/s, 56.4\% SLO) and beats the static default (54.2\%) and the
independent auto-tuner (53.4\%), with 98\% regime-detection accuracy.

\subsection{Ablation Study}
\label{sec:ablation}

Freezing each coordinated component at its global default, on the non-stationary
workload, isolates its contribution (Table~\ref{tab:ablation}). Removing batch
adaptation costs the most (60.4 $\rightarrow$ 54.6 tok/s), followed by eviction
adaptation; disabling cache-affinity routing is neutral here because the
non-stationary stream contains no cache-heavy regime --- exactly the expected
behavior, confirming the knob is inert when irrelevant rather than harmful.

\begin{table}[t]
\centering
\small
\caption{Ablation: CARL with each component frozen (non-stationary). Control-loop simulation.}
\label{tab:ablation}
\begin{tabular}{lrr}
\toprule
Variant & tok/s & SLO\% \\
\midrule
carl\_full              & 60.4 & 37.3 \\
\;\;-- no spec decode   & 58.0 & 35.5 \\
\;\;-- no batch adapt   & 54.6 & 34.7 \\
\;\;-- no eviction adapt& 55.4 & 35.2 \\
\;\;-- no cache affinity& 60.4 & 37.3 \\
\bottomrule
\end{tabular}
\end{table}

\section{Discussion}
\label{sec:discussion}

\paragraph{When CARL helps most.} The gains concentrate at regime transitions and
in regimes whose optimum is a knob \emph{combination} (e.g.\ batch: large batch
$+$ deep speculation). Where one knob dominates and a per-component tuner can
reach it, CARL's advantage shrinks to its warm start --- which, by design, is
already the expert baseline.

\paragraph{Limitations.} (1) The evaluation is a simulation; absolute numbers are
illustrative and a GPU wall-clock harness is the clear next step. (2) The
controller targets a single GPU; coordinating across a disaggregated or
multi-replica deployment adds cross-node state. (3) The explore/exploit trade-off
means early intervals may select sub-optimal arms; the warm start bounds this
downside but does not eliminate exploration cost, visible in the Thompson
variant's higher adaptation count and slightly lower reward. (4) The regime set
and arm sets are hand-designed; learning them is open.

\section{Related Work}
\label{sec:related}

CARL builds directly on continuous batching~\cite{yu2022orca}, paged and radix KV
management~\cite{kwon2023vllm, zheng2024sglang}, chunked
prefill~\cite{agrawal2024sarathiserve}, speculative
decoding~\cite{leviathan2023speculative}, and H2O eviction~\cite{zhang2023h2o},
treating each as a tunable subsystem rather than a fixed policy. Where
Clockwork~\cite{gujarati2020clockwork} delivers predictability bottom-up and
DistServe~\cite{zhong2024distserve} / AlpaServe~\cite{li2023alpaserve} optimize
placement and disaggregation, CARL is orthogonal: it adapts the joint runtime
configuration online. Our learning machinery is standard contextual
bandits~\cite{li2010contextual, agrawal2013thompson}; the novelty is the action
space (a joint serving configuration) and the per-regime structure.

\section{Conclusion}
\label{sec:conclusion}

Component-level optimization leaves joint, workload-dependent gains on the table.
CARL closes that gap with a single coordinated controller: a fused runtime-state
observer, a regime classifier, a joint configuration space, and a per-regime
contextual bandit that adapts the whole engine online and thread-safely. In a
control-loop evaluation over non-stationary and real LMSYS traffic, CARL tracks a
perfect-knowledge oracle and outperforms static and independently-tuned
baselines, adapting to regime shifts within one control interval. The natural
next step is a wall-clock GPU harness to convert these control-policy results
into measured end-to-end speedups.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
